---
title: "Portfolio"
author: "Athena"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(httr)
library(jsonlite)
library(rvest)
library(lubridate)
library(dplyr)
library(usmap)
library(ggplot2)
library(tidyverse)
library(choroplethr)
library(choroplethrMaps)
library(shiny)
library(leaflet)    
library(shiny)
library(shinydashboard)
library(geojsonio)
library(sp)
```

For this project I'm making a shiny page focusing on Head Start data 

Using Web scrapping to get a table of poverty line data overtime, which will be used on my shiny page 

```{r}
pov_link = "http://aspe.hhs.gov/topics/poverty-economic-mobility/poverty-guidelines/prior-hhs-poverty-guidelines-federal-register-references"

pov_page = read_html(pov_link)
table_poverty_line = pov_page %>% html_nodes("table.footable.table-align-left")%>%
  html_table() %>%.[[1]]

```

Using webscraping to get a quote about Head Start I will use in my shiny page 

```{r}
HS_quote_link = "https://eclkc.ohs.acf.hhs.gov/programs/article/head-start-approach"
HS_quote_page = read_html(HS_quote_link) %>% html_nodes("p:nth-child(1) , p:nth-child(2)") %>% html_text()
HS_quote <-  HS_quote_page[-c(1, 2)] 

quote <- paste( unlist(HS_quote), collapse=' ')
print(quote)
```



Don't forget to set your path! 

```{r}
setwd('C:/Users/athen/OneDrive/Documents/GitHub/data-analysis-portfolio/Shiny_R_Project')
head_start_data = read.csv("Head Start enrollment by age group.csv", header = TRUE)
population_data = read.csv("Child population by single age.csv", header = TRUE)
location_data = read.csv("Head Start Center Locations.csv", header = TRUE)
working = read.csv("low_income_working.csv", header = TRUE)
not_working = read.csv("children_no_working_parents.csv", header = TRUE)
```


For this analysis I'm only going to focus on kids under four because these are the ones primarily in Head Start 
```{r}
population_data <- subset(population_data, Single.Age != "") 

population_data$Single.Age[population_data$Single.Age == '1' | population_data$Single.Age == '<1'| population_data$Single.Age == '2'] <- '<3'

```

```{r}
colnames(population_data)[colnames(population_data) == 'Single.Age'] <- "Age.group"
colnames(population_data)[colnames(population_data) == 'Data'] <- "Population"
colnames(head_start_data)[colnames(head_start_data) == 'Data'] <- "Headstart"
population_data <- subset(population_data, select = -DataFormat)
head_start_data <- subset(head_start_data, select = -DataFormat)
```

I'm merging population and Head Start data based on age and state, and year so I can calculate a percentage 


```{r}
merged = merge(population_data, head_start_data, by =c('Age.group', 'Location', 'TimeFrame', 'LocationType'))
colnames(merged)[colnames(merged) == 'Location'] <- "state"
merged['name']<- merged['state']
merged$state= state.abb[match(merged$state,state.name)]


```

The location data also included US territories,so I'm removing them and creating a percentage.



```{r}
state_list <- as.list(unique(merged$state))
location_data <- subset(location_data, state %in% state_list) 
merged$year <- as.numeric(merged$TimeFrame)
merged["value"]= 100 * (as.numeric(merged$Headstart)/as.numeric(merged$Population))

```

I also want to create a data set which included percent of low income children who are in head start. Because the datasets I had were aggregated based on whether parents were working 
```{r}
head_start_data_total <- head_start_data[head_start_data$Age.group == 'Total',]
```

Cleaning the data for working and not working families add adding them 
```{r}
colnames(working)[colnames(working) == 'Data'] <- "Working"
working <- working[working$Age.group == "less than 6",]

colnames(not_working)[colnames(not_working) == 'Data'] <- "Not Working"
lowincome = merge(working, not_working, by =c('Location', 'TimeFrame', 'LocationType'))

lowincome = merge(lowincome, head_start_data_total, by =c('Location', 'TimeFrame', 'LocationType'))

```

There were two NAs some of the results for DC was missing for two years
DC also doesn't appear on the map, so I'm dropping it by filtering by state
```{r}
lowincome$value<- 100* as.numeric(lowincome$Headstart)/(as.numeric(lowincome$Working) + as.numeric(lowincome$`Not Working`))
lowincome$name <- lowincome$Location
lowincome$year<- as.numeric(lowincome$TimeFrame)
lowincome <- lowincome[lowincome$LocationType == "State",]
```



Here I'm matching latitude and longitude points and county data to map states enrollment in Head Start and map Head Start locations

```{r}
head_start_lonlat <- data.frame(lat = location_data$latitude, lon = location_data$longitude, state = location_data$state)
transformed_data <- usmap_transform(head_start_lonlat)

```

```{r}
location_data$lat <- location_data$latitude
location_data$lon <- location_data$longitude
transformed_data_state<- usmap_transform(location_data)

```


I'm matching counties with Fip codes to be able to map on a county level 
```{r}
Fips <- location_data %>% 
    group_by(state) %>% 
    summarise(region = list(fips(state = first(state), county = county))) %>%
    unnest(c(region))

Fips <- transform(Fips, values = ave(seq(nrow(Fips)), region, FUN=length))
Fips$fips <- as.numeric(Fips$region)
Unique_Fips <- Fips[!duplicated(Fips["fips"]),]
us <- usmap::us_map()
```

I'm also loading geo-data to make an interactive leaflet map

```{r}
states <- geojsonio::geojson_read("https://rstudio.github.io/leaflet/json/us-states.geojson", what = "sp")
```





Run this to create the shiny page 
```{r}
choiceName <- c("Yes", "No")

ui <- fluidPage(
   fluidRow( 
  titlePanel(
    h1("Learn More about Enrolled in Head start", align = "center", style = "font-weight: 500; color: white; background-color: #000080"))),
   fluidRow( 
      tabsetPanel(
        id = "tabset",
        tabPanel("Introduction", 
                 
   h2("Introduction", align = "center", style = "font-weight: 400"),
  fluidRow( column(12, offset = 2, mainPanel(p(quote)))),
   fluidRow( column(12, offset= 3,
  mainPanel(p("-“Head Start Approach | ECLKC.” 2022. August 9, 2022. https://eclkc.ohs.acf.hhs.gov/programs/article/head-start-approach")))),
  fluidRow(column(12,align="center", plotOutput("photo"))),
  fluidRow( column(12, offset = 2, mainPanel(p("Photograph, Lady Bird Johnson Visiting a Project Head Start Classroom,
March 19, 1966
Johnson White House Photographs
National Archives, White House Photo Office Collection (National Archives Identifier 596401)"))))),
        tabPanel("Nation Wide", fluidRow(
  column(3, offset = 2, selectInput("age", label = "Pick an Age Group", choices = unique(merged$Age.group))),
  column(4, sliderInput("year", "Pick a Year", value = 2019, min = 1990, max = 2019, sep = ""))),
    titlePanel(
    h2(textOutput("description"), align = "center", style = "font-weight: 300")),
  fluidPage(leafletOutput("mymap")),
  titlePanel(
    h2(textOutput("map2_des"), align = "center", style = "font-weight: 300")),
  fluidPage(leafletOutput("map_income")),
  fluidRow( column(12, offset = 2, mainPanel(p("Population Division, U.S. Census Bureau. 2023. “Office of Head Start (OHS). 2022. “Head Start Enrollment by Age Group, Annual Program Information Report (PIR).” Kids Count Data Center. https://datacenter.aecf.org/data/tables/9786-head-start-enrollment-by-age-group?loc=1&loct=2#detailed/2/2-53/false")))),
    fluidRow( column(12, offset = 2, mainPanel(p("Population Reference Bureau, U.S. Census Bureau. 2022. “The Share of Children under Age 6 Whose Resident Parents Are Not in the Civilian Labor Force.” Kids Count Data Center. https://datacenter.aecf.org/data/tables/5056-children-under-age-6-with-no-parent-in-the-labor-force?loc=1&loct=2#detailed/2/2-53/false/")))),
  fluidRow( column(12, offset = 2, mainPanel(p("Population Reference Bureau, U.S. Census Bureau. 2022. 2022b. “Children in Low-Income Working Families by Age Group in United States.” Kids Count Data Center. https://datacenter.aecf.org/data/tables/5048-children-in-low-income-working-families-by-age-group?loc=1&loct=1#detailed/1/any/false/2048,1729,37,871,870,573,869,36,868,867/34,35,36/11455,11456.")))),
  fluidRow( column(12, offset = 2, mainPanel(p("Population Division, U.S. Census Bureau. 2023. “Child Population by Single Age.” Kids Count Data Center. https://datacenter.aecf.org/data/tables/100-child-population-by-single-age?loc=1&loct=2#detailed/2/2-53/false/"))))),
        tabPanel("Head Start State Locations",  
                 h2(textOutput("states_name"), align = "center", style = "font-weight: 300"),
                 fluidRow( column(3, offset = 4, selectInput("state", label = "Pick a State to learn about?", choices = sort(unique(location_data$state))))),
    fluidRow(column(6, offset = 4, checkboxInput("location_points", "Do you want to overlay state map with Head Start Location Cordinates?", value = F))),
  fluidRow( column(12, offset = 1,
  plotOutput("state_plot", width = 1000))),
  fluidRow( column(12, offset= 3,
  mainPanel(p("“Head Start Center Location Datasets | ECLKC.” n.d. Accessed October 2, 2023. https://eclkc.ohs.acf.hhs.gov/about-us/article/head-start-center-location-datasets."))))
                 ),
        tabPanel("Poverty Data", h2("Poverty Line Overtime", align = "center", style = "font-weight: 300"),
        fluidRow( column(12, offset = 2, mainPanel(p("Head Start elligibilty is based on the federal poverty line, this table shows what the federal poverty line overtime")))),
        fluidRow(column(offset = 3, 
            dataTableOutput(outputId = "table"), width = 7),
            
    )))),)

server <- function(input, output, session) {
  output$photo <- renderImage({
    filename <- normalizePath("C:/Users/athen/OneDrive/Documents/GitHub/data-analysis-portfolio/Shiny_R_Project/Images/head_start_picture.jpeg")
        list(src = filename)
  }, deleteFile = FALSE)
  
output$description <- renderText({ 
    text_year <- input$year
    age_range <- input$age
      paste(text_year, " Percent of Children ", age_range, " Years Old Enrolled in Head Start" )
  })

output$map2_des <- renderText({ 
  if(input$year > 2008){ 
    text_year <- input$year} else {
    text_year <- 2008 }
    paste(text_year, "Percent of Low Income Children Under Six in Head Start")
  })


output$states_name <- renderText({ 
    text_state <- input$state
    text_state <- state.name[match(text_state,state.abb)]
      paste("2023 Head Start Locations in ", text_state)
  })
    output$table <- renderDataTable({table_poverty_line}, 
        options = list(scrollX = FALSE))

output$static <- renderTable(table_poverty_line)
output$state_plot <- renderPlot({{
  if(input$location_points){ 
    transformed_data_state <- transformed_data[transformed_data$state == input$state ,]
  plot_usmap(regions = "counties", data = Unique_Fips, color = "black", include = c(input$state)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))+
    labs(title = "Number of Head Start Locations Per County") + 
    theme(panel.background = element_rect(colour = "black"))+
    ggplot2::scale_fill_distiller(palette = "YlGnBu", name = "Number of Head Start Locations \n (Scale Changes Based on State)", label = scales::comma) + theme(legend.position = "right") +
  geom_point(data = transformed_data_state, 
             aes(x = x, y = y), 
             color = "black",
             fill = "#FFBF1F",
             size = 1.5,
             shape = 21)
    
  }else{
    
    transformed_data_state <- transformed_data[transformed_data$state == input$state ,]
  plot_usmap(regions = "counties", data = Unique_Fips, color = "black", include = c(input$state)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))+
    labs(title = "Number of Head Start Locations Per County") + 
    theme(panel.background = element_rect(colour = "black"))+
    ggplot2::scale_fill_distiller(palette = "YlGnBu", name = "Number of Head Start Locations \n (Scale Changes Based on State)", label = scales::comma) + theme(legend.position = "right")
    }}
})

output$mymap <- renderLeaflet({{
   reactive_df <- merged[merged$Age.group == input$age & merged$year == input$year, ]
  spacial_maping <- sp::merge(x= states, y=reactive_df, by ='name', duplicateGeoms = TRUE)
pal <- colorNumeric("Spectral", domain = c(0,45), reverse = TRUE, na.color = NA)
labels <- sprintf(
  "<strong>%s</strong><br/>%g%%",
  spacial_maping$name, round(spacial_maping$value, digits = 2)
) %>% lapply(htmltools::HTML)

leaflet(spacial_maping) %>%
  setView(-96, 37.8, 4) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(value),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = pal, values = ~value, opacity = 0.7, title = "Percent of Children </br> in Head Start",
    position = "bottomright") 
  }})

output$map_income <- renderLeaflet({{
  if(input$year > 2008){ 
   reactive_df <- lowincome[lowincome$year == input$year, ]}
  else {reactive_df <-lowincome[lowincome$year == 2008, ]}
  spacial_maping <- sp::merge(x= states, y=reactive_df, by ='name', duplicateGeoms = TRUE)
pal <- colorNumeric("RdYlBu", domain = c(0,35), reverse = TRUE, na.color = NA)
labels <- sprintf(
  "<strong>%s</strong><br/>%g%%",
  spacial_maping$name, round(spacial_maping$value, digits = 2)
) %>% lapply(htmltools::HTML)

leaflet(spacial_maping) %>%
  setView(-96, 37.8, 4) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(value),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = pal, values = ~value, opacity = 0.7, title = "Percent of Low Income </br> Children Under Six </br> in Head Start",
    position = "bottomright") 
  }})

}
shinyApp(ui, server)
```


